Best Practices in Mixture Modeling using Free Open Source Software

### 2022-09-14 manuscript.Rmd
Update 14 Septemeber
In the manuscript.Rmd:

Added Shapiro-Wilks test when discussing Spurk.
Removed some obsolete comments (we still have backup just in case).
Wrote about alternative model parametrizations in LCA, referring back to GRoLTS.
Under Model Fit Indices, I wrote about -2*LL, mentioned we are working on an implementation of the lazy bootstrap.
Finished the Classification Diagnostics and Interpreting Class Solution sections.

In Daniel_LCA:

Finished annotating the code. We can include this section in the manuscript in the next commit.

abstract: |
  Latent class analysis is a popular technique for identifying groups in data based
  on a parametric model. Examples of this technique are known as mixture models, latent profile
  analysis, latent class analysis, growth mixture modeling, and latent class
  growth analysis. Despite the popularity of this technique, there is limited
  guidance with respect to best practices in estimating and reporting mixture
  models. Moreover, although user-friendly interfaces for advanced mixture
  modeling have long been available in commercial software packages, open
  source alternatives have remained somewhat inaccessible. This tutorial
  describes best practices for the estimation and reporting of latent class analysis,
  using free and open source software in R. To this end, this tutorial
  introduces new functionality for estimating and reporting mixture
  models in the `tidySEM` R-package. These functions rely on estimation using
  the OpenMx R-package.

@@

```{r echo = FALSE}
library(tidySEM)
library(ggplot2)
set.seed(1)
n = 100
C <- sample(c("Man", "Woman"), n, replace = TRUE)
means <- c(Man = 9, Woman = 7)
X <- rnorm(n, mean = means[C], sd = 1)
X <- round(X*2)/2
obsparam <- tapply(X, C, mean)
```

@@

When the focus is on each individual's estimated class membership,
latent class analysis can be thought of as a type of clustering algorithm.
In line with this perspective, mixture modeling is sometimes described as "model-based
clustering" [@hennig_handbook_2015; @scrucca_mclust_2016].
Many clustering algorithms apply some recursive splitting algorithm to the data.
By contrast "model-based" clustering refers to the fact that latent class analysis 
estimates cluster membership based on a parametric model.
Specifically, the posterior class probability that an individual belongs to a latent class
can be computed from the likelihood of that individual's observed data under given the class-specific model.

Finally, in the context of machine learning, latent class analysis can be considered as an
*unsupervised classification* problem [@figueiredo_unsupervised_2002].
The term *unsupervised* refers to the fact that the outcome variable, true class membership,
is not known, and the term *classification* refers to the fact that the algorithm is predicting a categorical outcome (class membership).

@@

The technique known as *latent profile analysis (LPA)* is a special case of the mixture model,
which assumes conditional independence of the indicators.
Conditional independence means that, after class membership is accounted for,
the covariances/correlations between indicators are assumed to be zero.
This can be conceived of as a restricted mixture model with covariances fixed to zero.
In some cases, such constraints will be inappropriate;
for example, when the cohesion between indicators is expected to differ between classes.
As an example, a mixture model analysis of ocean plastic particles found two classes of particles based on length and width:
A class of smaller particles with a high correlation between length and width, meaning that these particles were approximately round or square in shape, and a class of larger particles with a low correlation between length and width, meaning that these particles were heterogenous in shape.
From a theoretical perspective, this makes sense, because the smaller particles have been ground down to a more uniform shape by the elements.

@@

This paper will primarily focus on mixture models and special cases thereof,
although most of the suggested guidelines are applicable to latent class analyses.

<!-- DA: Do we still need the following comment? In my opinion the included text is sufficient. -->
<!-- If a multivariate mixture estimation is constrained so that measures must be uncorrelated within each distribution it is termed latent profile analysis. Modified to handle discrete data, this constrained analysis is known as LCA. Discrete latent trait models further constrain the classes to form from segments of a single dimension: essentially allocating members to classes on that dimension: an example would be assigning cases to social classes on a dimension of ability or merit. -->

<!-- Because LPA is model-based, a number of different model parameterizations can be -->
<!-- estimated. These models differ in terms of whether--and how--parameters are -->
<!-- estimated across the profiles. These parameters are the means for the different -->
<!-- profiles, which, in this approach, always are estimated freely across the -->
<!-- profiles; the variances for the variables used to create the profiles, which can -->
<!-- be estimated freely or can be estimated to be the same, or equal, across -->
<!-- profiles; and the covariances of the variables used to create the profiles, -->
<!-- which can be freely-estimated, estimated to be equal, or fixed to be zero. -->

<!-- In general, as more parameters are estimated (i.e., those that are fixed to zero -->
<!-- are estimated as being equal across profiles; or those estimated as being equal -->
<!-- across profiles are freely-estimated across them), the model becomes more -->
<!-- complex; the model may fit better, but also be overfit, meaning that the -->
<!-- profiles identified may be challenging to replicate with another, separate data -->
<!-- set. Even still, flexibility in terms of which models can be estimated also has -->
<!-- affordances. For example, the varying means, equal variances, and covariances -->
<!-- fixed to 0. A researcher might choose this model specification if she wants to -->
<!-- model the variables to be used to create profiles that are independent. This -->
<!-- model is very simple, as no covariances are estimated and the variances are -->
<!-- estimated to be the same across profiles. As we estimate more parameters (and -->
<!-- decrease the degrees of freedom), we are more likely to fit the data, but less -->
<!-- likely to be able to replicate the model with a second set of data. As we -->
<!-- progress toward more complex models (with increasingly complex -->
<!-- parameterization), then we are more likely to fit the data better. -->
<!-- In all, this flexibility associated with LPA also has a more pragmatic cost. It -->
<!-- can be very difficult to efficiently---and reproducibly---estimate and compare a -->
<!-- number of models. This cost means that analysts may focus on a subset of models. -->
<!-- It may also mean that unintentional errors are introduced through copying and -->
<!-- pasting the output of model estimations. Part of why we developed tidyLPA was to -->
<!-- make the model estimation process more efficient and more reproducible. As we -->
<!-- describe later, we do this using R, which has a number of benefits, including -->
<!-- being open-source, powerful (in terms of pre-processing and using the output of -->
<!-- models), and increasingly widely-used in the psychological sciences. We also -->
<!-- provide an interface to perhaps the most widely-used software for LPA, MPlus. In -->
<!-- this way, tidyLPA can both allow researchers to estimate models they already -->
<!-- estimate using MPlus more efficiently and reproducibly, and, for those without -->
<!-- access to this software, to do so entirely using R and the open-source R package -->
<!-- mclust. -->

@@

@@

LCA is also appropriate when we wish to identify indicators that capture classes well.
High quality indicators are strongly related to the latent variable
and lead to good class separation.
This relationship of high quality indicators to the latent variable
is reflected in very high or very low conditional response probabilities.
As a rule of thumb, high indicator quality can be indicated 
by conditional response probabilities above .9 and below .1. 
Moderate indicator quality can be indicated by probabilities above .8 and below .2,
and low indicator quality would be above .7 and below .3.
For a simulation study exploring the effects of indicator quality on LCA, see @geiser_is_2014.
Therefore, one could use conditional response probabilities for each item to assess its quality
with regards to how well it helps separate the latent classes.
From this, a theory about the selection of indicators could be informed.

@@

Finally, LCA can be used to deal with data which violate certain assumptions.
For instance, nested data usually imply a violation of the independence assumption.
In this case, an extension of LCA called multilevel LCA can be used (e.g. @henry_multilevel_2010).
As discussed in the shoe size example, LCA can deal with violations of normality.
In fact, LCA assumes the population distribution in a non-normal mixture of $K$ normal distributions,
and LCA can discover the value of $K$, i.e. generate a $K$-class solution.

<!-- check with Caspar if there are other assumptions and how to expand this paragraph -->

<!-- DA: An assumption LCA has is that of conditional or local independence. -->
<!-- LCA assumes conditional independence of indicators. -->
<!-- This means class membership explains all the relationships between the indicators, -->
<!-- and when the latent class is controlled for, the indicators become unrelated [@masyn_latent_2013] -->

<!-- OLD: -->
<!-- CJ: Include these use cases too: -->
<!-- Descriptive analysis; identify few prototypes based on many variables, e.g. DJA science funding -->
<!-- Inductively identify number of latent classes: I know it's a group, but not how many groups -->
<!-- Classify individuals, e.g. welzijnsmeter -->
<!-- Identify items (observed characteristics) that -->
<!-- indicate these classes well -->
<!-- Predict class membership from covariates -->
<!-- Predict outcomes from class membership -->
<!-- Dimension reduction -->
<!-- Deal with violation of assumptions, e.g. when data are clearly not normally distributed -->

@@

```{r checkest}
tab_check <- data.frame(Item = c(
  "Examine observed data",
  "Handling missing data",
  "Alternative model specifications",
  "Starting values",
  "Algorithm"
))
tab_check$`#` = paste0(1:length(tab_check$Item), ".")
kbl(tab_check[, c("#", "Item")])
```

@@

Examining observed data is essential for any analysis
as it may reveal patterns and violations of assumptions that had not been considered prior to data collection.
Pay special attention to level of measurement of the indicators.
Finite Gaussian mixture models (including LPA) are strictly speaking only suitable for continuous variables.
Indicators with an ordinal level of measurement are likely to violate the assumption of within-class normal distributions of mixture models [see @vermunt_k-means_2011].
Personal experience consulting on latent class analyses and moderating the `tidyLPA` Google group
suggest that the misapplication of mixture models to ordinal (e.g., Likert-type) indicators is the most common source of user error.
Whereas it has been argued that some parametric methods are robust when scales with 7+ indicators are treated as continuous [e.g., @norman_likert_2010],
this certainly does not imply that all methods are.
It is certainly unlikely that such ordinal variables can be treated as a *mixture* of multiple normal distributions.
The problem becomes egregious when the number of classes estimated equals or exceeds the number of categories; 
in this case, each class-specific mean could describe a single response category, 
and a class-specific variance component would be nonsensical.
In sum, Likert-type scales are rarely suitable for mixture modeling;
latent class analysis with ordinal indicators is more appropriate.

Relatedly, a recent publication claimed that an assumption of mixture models is that observed indicators are normally distributed [@spurk_latent_2020].
This is incorrect.
When the number of classes is greater than one, mixture models assume that the observed indicators are a mixture of multiple (multivariate) normal distributions.
In our shoe size example, it can be seen that the population distribution comprised of two normal distributions
is bimodal and hence not normal. 
Yet, this is a prototypical example of a mixed population distribution where LCA can discover latent groups.
If the population distribution were instead normal, there would be no classes to extract as the whole population would belong to a single class.

@@

<!-- DA: Does this following part about conditional item probabilities and inadmissible solutions belong here?  It's about model interpretation.-->

An important outcome of LCA are conditional item probabilities,
also known as class-specific item probabilities [@masyn_latent_2013], 
conditional response or conditional solution probabilities [@geiser_data_2012].
They indicate the probability of an item being endorsed
given that the observation belongs to a particular latent class.
Conditional item probabilities can be obtained using `tidySEM::table_prob()`
If a particular item is endorsed by two or more classes at markedly different rates,
it is said to discriminate well between the classes and is consequently considered a good indicator.
Classes are considered highly homogeneous with respect to an item when for a particular item
there is a distinct difference in conditional item probabilities for two or more classes.
For instance, if an item is endorsed below 30% for one class and above 70% for another class,
the classes have high homogeneity with respect to this item [@masyn_latent_2013].
Conditional item probabilities are the analogue of mean and standard deviation
when the indicators are binary/ordinal.

@@

While we concur that investigating missingness is due dilligence,
it is important to emphasize that missingness is adequately handled by default in many software packages for latent class analyses, such as `OpenMx` (and e.g., Mplus).
These packages use Full Information Maximum Likelihood (FIML) estimation,
which makes use of all available information without imputing missing values.
FIML is a best-practice solution for handling missing data; on par with multiple imputation [@lee_comparison_2021].
FIML estimation assumes that missingness is either MCAR or MAR.
Thus, one would typically proceed with FIML regardless of the outcome of an MCAR test.
Although FIML does not, by default, handle missingness in exogenous variables - all indicator variables in latent class analysis are endogenous, so this is not a concern.

@@

The different types of latent class models have different parameters.
For example, mixture models and latent profile analyses typically have class-specific means, variances, and covariances.
Latent growth analyses have the same parameters, but with respect to the latent growth variables.
Latent class analyses with ordinal indicators have thresholds.
All of these parameters can be freely estimated across classes, constrained to be equal across classes,
or fixed to a certain value (e.g., to zero).
The total number of parameters thus scales with the number of estimated classes.
Consequently, latent class analyses have a potentially very high number of parameters.
As any of these parameters could be misspecified,
it is important to consider alternative model specifications.
However, alternative model specifications may be approached differently depending on whether an analysis is data driven (exploratory), or theoretically driven (confirmatory).
This distinction has remained underemphasized in prior writing.

@@

Confirmatory analyses typically require less comprehensive alternative model specifications.
For example, in the context of preregistered analyses, the main models of interest may have been specified a priori.
Therefore, it may not be necessary to test different model specifications when using the confirmatory approach.

In order to aid researchers working with latent trajectory models,
@van_de_schoot_grolts-checklist_2017 developed a protocol called Guidelines for Reporting on Latent Trajectory Studies (GRoLTS).
Throughout this paper, we adapt those GRoLTS protocols that are applicable to latent class analysis.
One of the guidelines (namely, 6a) refers to considering alternative model specifications.
In LCA, means, variances, and covariances between the indicators 
can either be constrained across classes, or estimated freely.
Researchers using LCA should transparently report their chosen parametrization 
as well as discuss different parametrizations that were tested in the process.

<!-- DA: - would researchers ideally anticipate the parameterization of the within-class var/cov structure? How does preregistration connect to this?  -->

```{r}
grolts <- structure(list(Number = c("1.", "2.", "3a.", "3b.", "3c.", "4.", 
"5.", "6a.", "6b.", "7.", "8.", "9.", "10.", "11.", "12.", "13.", 
"14a.", "14b.", "14c.", "15.", "16."), Item = c("Is the metric of time used in the statistical model reported?", 
"Is information presented about the mean and variance of time within a wave?", 
"Is the missing data mechanism reported?", "Is a description provided of what variables are related to attrition/missing data?", 
"Is a description provided of how missing data in the analyses were dealt with?", 
"Is information about the distribution of the observed variables included?", 
"Is the software mentioned?", "Are alternative specifications of within-class heterogeneity considered (e.g., LGCA vs. LGMM) and clearly documented? If not, was sufficient justification provided as to eliminate certain specifications from consideration?", 
"Are alternative specifications of the between-class differences in variance–covariance matrix structure considered and clearly documented? If not, was sufficient justification provided as to eliminate certain specifications from consideration?", 
"Are alternative shape/functional forms of the trajectories described?", 
"If covariates have been used, can analyses still be replicated?", 
"Is information reported about the number of random start values and final iterations included?", 
"Are the model comparison (and selection) tools described from a statistical perspective?", 
"Are the total number of fitted models reported, including a one-class solution?", 
"Are the number of cases per class reported for each model (absolute sample size, or proportion)?", 
"If classification of cases in a trajectory is the goal, is entropy reported?", 
"Is a plot included with the estimated mean trajectories of the final solution?", 
"Are plots included with the estimated mean trajectories for each model?", 
"Is a plot included of the combination of estimated means of the final model and the observed individual trajectories split out for each latent class?", 
"Are characteristics of the final class solution numerically described (i.e., means, SD/SE, n, CI, etc.)?", 
"Are the syntax files available (either in the appendix, supplementary materials, or from the authors)?"
), relevant = c(FALSE, FALSE, TRUE, TRUE, TRUE, TRUE, TRUE, TRUE, 
TRUE, FALSE, TRUE, TRUE, TRUE, TRUE, TRUE, TRUE, TRUE, TRUE, 
TRUE, TRUE, TRUE)), row.names = c(NA, -21L), class = "data.frame")
```

@@

<!-- 3. Packages -->
<!-- 3.1 R-packages -->
<!-- LPA models can be fit in R (version 4.0.3; R-core team, 2020), running in Rstudio (version -->
<!-- 1.3.1093 used here). There are many R-packages that offer some form of latent class/mixture -->
<!-- analytic functionality. However, the majority of packages are focused on analyses with discrete -->
<!-- indicator variables (e.g., ‘poLCA’, Linzer & Lewis, 2011; ‘e1071::lca’, Meyer et al., 2019) -->
<!-- or require a lot of coding to define the required models (e.g., openMX, Neale et al., 2016). -->
<!-- For the current tutorial, two packages were selected that work with continuous indicator -->
<!-- variables and that require limited user coding. ‘mclust’ package (Scrucca et al., 2016) is the -->
<!-- first package that will be illustrated. This package is a specialist tool and allows for a wide -->
<!-- variety of model configurations to be estimated; as such it offers much more functionality -->
<!-- than most researchers will likely need. Therefore, the current tutorial focuses on a limited -->
<!-- range of relatively simple LPA model configurations that are commonly encountered in the -->
<!-- literature. Conveniently, Rosenberger et al., (2018) recently developed the package ‘tidyLPA’ -->
<!-- which can be used as a relatively easy to use front-end for estimating common LPA models -->
<!-- with ‘mclust’, basically streamlining some of the in- and output functionality of ‘mclust’. -->

@@

<!-- We assume that the population consists of several groups each characterized by its own normal distribution.  -->
<!-- When two or more normal distributions with different means and standard deviations comprise a mixed population distribution,  -->
<!-- this combined distribution is no longer normal.  -->
<!-- For example, you can imagine a bimodal distribution when dealing with two latent classes.  -->
<!-- If this were not the case and our mixed population distribution were normal,  -->
<!-- there would be no latent classes to extract as the population would consists of a single class only.  -->
<!-- For this reason, multivariate normality will always be violated when the population is comprised of several latent groups.  -->
<!-- Secondly, the claim that maximum likelihood with robust standard errors should be preferred  -->
<!-- implicitly implies that only software packages with this estimation technique can be used.  -->

@@

<!-- The EM algorithm needs starting values for each parameter it wants to estimate.  -->
<!-- If it were provided a single starting value for each parameter,  -->
<!-- the EM optimizer would be at risk of finding a local optimum instead of the global optimum.  -->
<!-- In other words, it would converge on a suboptimal solution resulting in biased parameter and model fit estimates.  -->
<!-- To prevent this, we should give the algorithm a sufficiently large set of random starting values for each parameter -->
<!-- in order to maximize our chances of finding the deepest valley that is the global optimum (McLachlan & Peel, 2004).  -->
<!-- Hipp and Bauer (2006) recommend using minimally 50-100 sets of random starting values. -->
<!-- <!-- add more on this --> -->
<!-- Also, we should ensure the number of initial stage iterations is large for optimization to be successful (Geiser, 2012).  -->

<!-- Furthermore, the choice of the starting values can impact the speed of the algorithm's convergence -->
<!-- or in case of having poor initial values the algorithm might diverge altogether (McLachlan & Peel, 2004). -->
<!-- The key is to use a large number of starting value sets.  -->
<!-- Geiser (2013) recommends at least 500 sets of initial values in the first optimization step for simple models, -->
<!-- and a greater number of sets for more complex models.  -->
<!-- A large number of starting values ensures that we find the true maximum and estimate the model parameters and fit accurately. -->

@@

### Model fit indices

@@

An alternative is using the bootstrapped likelihood ratio test
which can be run using `tidySEM::BLRT()`.
Currently, this test is computationally expensive and can be slow on most computers.
The development of a faster version of this test is well under way.

@@

Three important classification diagnostics provided by `tidySEM` are are entropy, 
the range of the posterior classification probabilities by most likely class membership,
and the range of the sample size.

>>>>

abstract: |
  Latent class analysis is a popular technique for identifying groups in data based
  on a parametric model. Examples of this technique are known as mixture models, latent profile
  analysis, latent class analysis, growth mixture modeling, and latent class
  growth analysis. Despite the popularity of this technique, there is limited
  guidance with respect to best practices in estimating and reporting mixture
  models. Moreover, although user-friendly interfaces for advanced mixture
  modeling have long been available in commercial software packages, open
  source alternatives have remained somewhat inaccessible. This tutorial
  describes best practices for the estimation and reporting of latent class analysis,
  using free and open source software in R. To this end, this tutorial
  introduces new functionality for estimating and reporting mixture
  models in the `tidySEM` R-package. These functions rely on estimation using
  the `OpenMx` R-package.

@@

```{r echo = FALSE}
library(tidySEM)
library(ggplot2)
set.seed(1)
n = 100
C <- sample(c("Man", "Woman"), n, replace = TRUE)
means <- c(Man = 9, Woman = 7)
X <- rnorm(n, mean = means[C], sd = 1)
X <- round(X*2)/2
obsparam <- tapply(X, C, mean)
shoesize_shapiro <- shapiro.test(X) # Shapiro-Wilk test of normality
```

@@

When the focus is on each individual's estimated class membership,
latent class analysis can be thought of as a type of clustering algorithm.
In line with this perspective, mixture modeling is sometimes described as "model-based
clustering" [@hennig_handbook_2015; @scrucca_mclust_2016].
Many clustering algorithms apply some recursive splitting algorithm to the data.
By contrast "model-based" clustering refers to the fact that LCA 
estimates cluster membership based on a parametric model.
Specifically, the posterior class probability that an individual belongs to a latent class
can be computed from the likelihood of that individual's observed data under given the class-specific model.

Finally, in the context of machine learning, LCA can be considered as an
*unsupervised classification* problem [@figueiredo_unsupervised_2002].
The term *unsupervised* refers to the fact that the outcome variable, true class membership,
is not known, and the term *classification* refers to the fact that the algorithm is predicting a categorical outcome -- class membership.

@@

The technique known as *latent profile analysis (LPA)* is a special case of the mixture model,
which assumes conditional independence of the indicators.
Conditional independence means that, after class membership is accounted for,
the covariances/correlations between indicators are assumed to be zero.
This can be conceived of as a restricted mixture model with covariances fixed to zero.
In some cases, such constraints will be inappropriate;
for example, when the cohesion between indicators is expected to differ between classes.
As an example, a mixture model analysis of ocean plastic particles found two classes of particles based on length and width:
a class of smaller particles with a high correlation between length and width, meaning that these particles were approximately round or square in shape, and a class of larger particles with a low correlation between length and width, meaning that these particles were heterogenous in shape.
From a theoretical perspective, this makes sense, because the smaller particles have been ground down to a more uniform shape by the elements.
<!-- DA: what is the source for plastic particles? -->

@@

This paper will primarily focus on mixture models and special cases thereof,
although most of the suggested guidelines are applicable to all latent class analyses.

@@

<!-- CJ: It's good but for practical purposes it's preferable -->
<!-- to use examples from Caspar's published papers. -->
<!-- Don't spend too much time on this, -->
<!-- Caspar will check if I can replace them with any of my papers. -->

@@

LCA is also appropriate when we wish to identify indicators that capture classes well.
High quality indicators are strongly related to the latent variable
and lead to good class separation.
This relationship of high quality indicators to the latent variable
is reflected in very high or very low conditional response probabilities.
<!-- As a rule of thumb, high indicator quality can be indicated  -->
<!-- by conditional response probabilities above .9 and below .1.  -->
<!-- Moderate indicator quality can be indicated by probabilities above .8 and below .2, -->
<!-- and low indicator quality would be above .7 and below .3. <- DA's writing -->
<!-- This quality rule of thumb is from @geiser_is_2014 where they state that -->
<!-- "Indicators in high quality conditions were generated to either have CRPs (of endorsing the second category) of 0.9 or 0.1, while moderate quality conditions had CRPs of 0.8 and 0.2, and low quality indicators were generated at 0.7 or 0.3, following Collins and Wugalter’s (1992) strong and weak measurement strength conditions." direct quote from @geiser_is_2014 -->
For a simulation study exploring the effects of indicator quality on LCA, see @geiser_is_2014.
Therefore, one could use conditional response probabilities for each item to assess its quality
with regards to how well it helps separate the latent classes.
From this, a theory about the selection of indicators could be informed.

@@

<!-- CJ: (on the next paragraph)  -->
<!-- the multilevel LCA addresses violations of the assumptions of LCA. -->
<!-- That is not what this paragraph is about.  -->
<!-- This paragraph is about motivations for using LCA. -->
<!-- One motivation is if your data violates multivariate normality -->
<!-- because violating multivariate normality means that you have multiple groups -->
<!-- and you can use LCA to extract those. That is the only assumption  -->
<!-- we have to discuss here, we just need to explain it. -->
Finally, LCA can be used to deal with data which violate certain assumptions.
As discussed in the shoe size example, LCA can deal with violations of normality.
In fact, LCA assumes the population distribution in a non-normal mixture of $K$ normal distributions,
and it can discover the value of $K$, i.e. generate a $K$-class solution.

<!-- CJ: Include these use cases too: -->
<!-- Descriptive analysis; identify few prototypes based on many variables, e.g. DJA science funding -->
<!-- Inductively identify number of latent classes: I know it's a group, but not how many groups -->
<!-- Classify individuals, e.g. welzijnsmeter -->
<!-- Identify items (observed characteristics) that -->
<!-- indicate these classes well -->
<!-- Predict class membership from covariates -->
<!-- Predict outcomes from class membership -->
<!-- Dimension reduction -->
<!-- Deal with violation of assumptions, e.g. when data are clearly not normally distributed -->

@@

```{r checkest, include=FALSE}
tab_check <- data.frame(Item = c(
  "Examine observed data",
  "Handling missing data",
  "Alternative model specifications",
  "Software",
  "Best practices in estimation",
  "Algorithm",
  "Class enumeration",
  "Model Fit Indices",
  "Classification Diagnostics",
  "Interpreting Class Solution",
  "Label switching",
  "Best Practices in Reporting",
  "Tutorial"

))
tab_check$`#` = paste0(1:length(tab_check$Item), ".")
kbl(tab_check[, c("#", "Item")])
```

@@

Examining observed data is essential for any analysis
as it may reveal patterns and violations of assumptions that had not been considered prior to data collection.
Special attention should be paid to level of measurement of the indicators.
Finite Gaussian mixture models (including LPA) are only suitable for continuous variables.
Indicators with an ordinal level of measurement are likely to violate the assumption of within-class normal distributions of mixture models [see @vermunt_k-means_2011].
Personal experience consulting on latent class analyses and moderating the `tidyLPA` Google group
suggest that the misapplication of mixture models to ordinal (e.g., Likert-type) indicators is the most common source of user error.
Whereas it has been argued that some parametric methods are robust when scales with 7+ indicators are treated as continuous [e.g., @norman_likert_2010],
this certainly does not imply that all methods are.
It is certainly unlikely that such ordinal variables can be treated as a *mixture* of multiple normal distributions.
The problem becomes egregious when the number of classes estimated equals or exceeds the number of categories; 
in this case, each class-specific mean could describe a single response category, 
and a class-specific variance component would be nonsensical.
In sum, Likert-type scales are rarely suitable for mixture modeling;
latent class analysis with ordinal indicators is more appropriate.

Relatedly, a recent publication claimed that an assumption of mixture models is that observed indicators are normally distributed [@spurk_latent_2020].
This is incorrect.
When the number of classes is greater than one, mixture models assume that the observed indicators are a mixture of multiple (multivariate) normal distributions.
In our shoe size example, it can be seen that the population distribution comprised of two normal distributions.
When examined visually, the population distribution is evidently bimodal. 
The Shapiro-Wilk normality test ($W = 0.971$, $p < 0.05$ ) 
rejects the null hypothesis that 
the sample comes from a normally distributed population.
Yet, this is a prototypical example of a mixed population distribution where LCA can discover latent groups.
If the population distribution were instead normal, 
there would be no classes to extract 
as the whole population would belong to a single class.

@@

@@

While we concur that investigating missingness is due dilligence,
it is important to emphasize that missingness is adequately handled by default in many software packages for latent class analyses, 
such as Mplus, and `OpenMx` which is the backend of `tidySEM`.
These packages use Full Information Maximum Likelihood (FIML) estimation,
which makes use of all available information without imputing missing values.
FIML is a best-practice solution for handling missing data; on par with multiple imputation [@lee_comparison_2021].
FIML estimation assumes that missingness is either MCAR or MAR.
Thus, one would typically proceed with FIML regardless of the outcome of an MCAR test.
Although FIML does not, by default, handle missingness in exogenous variables - all indicator variables in latent class analysis are endogenous, so this is not a concern.

@@

In order to aid researchers working with latent trajectory models,
@van_de_schoot_grolts-checklist_2017 developed a protocol called Guidelines 
for Reporting on Latent Trajectory Studies (GRoLTS).
Two of the GRoLTS guidelines (namely, 6a and 6b) refer to considering alternative model specifications.
Both are discussing specific cases in latent trajectory model specification. 
The first is about whether the variance of the growth parameter is estimated freely or fixed,
and the second is about whether conditional independence is assumed
as the researcher might also want to free-up the variance-covariance structure.
In LCA, there are also many different ways to specify the model.
Means, variances, and covariances between the indicators 
can either be constrained across classes, or estimated freely.
Researchers using LCA should transparently report their chosen parametrization 
as well as discuss different parametrizations that were tested in the process.

Different types of latent class models have different parameters.
For example, mixture models and latent profile analyses 
typically have class-specific means, variances, and covariances.
Latent growth analyses have the same parameters, but with respect to the latent growth variables.
Latent class analyses with ordinal indicators have thresholds.
All of these parameters can be freely estimated across classes, constrained to be equal across classes,
or fixed to a certain value (e.g., to zero).
The total number of parameters thus scales with the number of estimated classes.
Consequently, latent class analyses have a potentially very high number of parameters.
As any of these parameters could be misspecified,
it is important to consider alternative model specifications.
However, alternative model specifications may be approached differently depending on whether an analysis is data driven (exploratory), or theoretically driven (confirmatory).
This distinction has remained underemphasized in prior writing.

@@

Confirmatory analyses typically require less comprehensive alternative model specifications.
For example, in the context of preregistered analyses, 
the main models of interest may have been specified a priori. 
Even in case of confirmatory LCA, the theoretical model could be compared to a few others to contextualize it.

```{r, include = FALSE}
grolts <- structure(list(Number = c("1.", "2.", "3a.", "3b.", "3c.", "4.", 
"5.", "6a.", "6b.", "7.", "8.", "9.", "10.", "11.", "12.", "13.", 
"14a.", "14b.", "14c.", "15.", "16."), Item = c("Is the metric of time used in the statistical model reported?", 
"Is information presented about the mean and variance of time within a wave?", 
"Is the missing data mechanism reported?", "Is a description provided of what variables are related to attrition/missing data?", 
"Is a description provided of how missing data in the analyses were dealt with?", 
"Is information about the distribution of the observed variables included?", 
"Is the software mentioned?", "Are alternative specifications of within-class heterogeneity considered (e.g., LGCA vs. LGMM) and clearly documented? If not, was sufficient justification provided as to eliminate certain specifications from consideration?", 
"Are alternative specifications of the between-class differences in variance–covariance matrix structure considered and clearly documented? If not, was sufficient justification provided as to eliminate certain specifications from consideration?", 
"Are alternative shape/functional forms of the trajectories described?", 
"If covariates have been used, can analyses still be replicated?", 
"Is information reported about the number of random start values and final iterations included?", 
"Are the model comparison (and selection) tools described from a statistical perspective?", 
"Are the total number of fitted models reported, including a one-class solution?", 
"Are the number of cases per class reported for each model (absolute sample size, or proportion)?", 
"If classification of cases in a trajectory is the goal, is entropy reported?", 
"Is a plot included with the estimated mean trajectories of the final solution?", 
"Are plots included with the estimated mean trajectories for each model?", 
"Is a plot included of the combination of estimated means of the final model and the observed individual trajectories split out for each latent class?", 
"Are characteristics of the final class solution numerically described (i.e., means, SD/SE, n, CI, etc.)?", 
"Are the syntax files available (either in the appendix, supplementary materials, or from the authors)?"
), relevant = c(FALSE, FALSE, TRUE, TRUE, TRUE, TRUE, TRUE, TRUE, 
TRUE, FALSE, TRUE, TRUE, TRUE, TRUE, TRUE, TRUE, TRUE, TRUE, 
TRUE, TRUE, TRUE)), row.names = c(NA, -21L), class = "data.frame")
```

@@

<!-- 3. Packages -->
<!-- 3.1 R-packages -->
<!-- LPA models can be fit in R (version 4.0.3; R-core team, 2020), running in Rstudio (version -->
<!-- 1.3.1093 used here). There are many R-packages that offer some form of latent class/mixture -->
<!-- analytic functionality. However, the majority of packages is focused on analyses with discrete -->
<!-- indicator variables (e.g., ‘poLCA’, Linzer & Lewis, 2011; ‘e1071::lca’, Meyer et al., 2019) -->
<!-- or require a lot of coding to define the required models (e.g., openMX, Neale et al., 2016). -->
<!-- For the current tutorial, two packages were selected that work with continuous indicator -->
<!-- variables and that require limited user coding. ‘mclust’ package (Scrucca et al., 2016) is the -->
<!-- first package that will be illustrated. This package is a specialist tool and allows for a wide -->
<!-- variety of model configurations to be estimated; as such it offers much more functionality -->
<!-- than most researchers will likely need. Therefore, the current tutorial focuses on a limited -->
<!-- range of relatively simple LPA model configurations that are commonly encountered in the -->
<!-- literature. Conveniently, Rosenberger et al., (2018) recently developed the package ‘tidyLPA’ -->
<!-- which can be used as a relatively easy to use front-end for estimating common LPA models -->
<!-- with ‘mclust’, basically streamlining some of the in- and output functionality of ‘mclust’. -->

@@

<!-- We assume that the population consists of several groups each characterized by its own normal distribution. -->
<!-- When two or more normal distributions with different means and standard deviations comprise a mixed population distribution, -->
<!-- this combined distribution is no longer normal. -->
<!-- For example, you can imagine a bimodal distribution when dealing with two latent classes. -->
<!-- If this were not the case and our mixed population distribution were normal, -->
<!-- there would be no latent classes to extract  -->
<!-- as the population would consists of a single class only. -->
<!-- For this reason, multivariate normality will always be violated when the population is comprised of several latent groups. -->
<!-- Secondly, the claim that maximum likelihood with robust standard errors should be preferred -->
<!-- implicitly implies that only software packages with this estimation technique can be used. -->

@@

<!-- The EM algorithm needs starting values for each parameter it wants to estimate. -->
<!-- If it were provided a single starting value for each parameter, -->
<!-- the EM optimizer would be at risk of finding a local optimum instead of the global optimum. -->
<!-- In other words, it would converge on a suboptimal solution -->
<!-- resulting in biased parameter and model fit estimates. -->
<!-- To prevent this, we should give the algorithm a sufficiently large set of random starting values for each parameter -->
<!-- in order to maximize our chances of finding the deepest valley that is the global optimum (McLachlan & Peel, 2004). -->
<!-- Hipp and Bauer (2006) recommend using minimally 50-100 sets of random starting values. -->

<!-- Also, we should ensure the number of initial stage iterations is large for optimization to be successful (Geiser, 2012). -->

<!-- Furthermore, the choice of the starting values  -->
<!-- can impact the speed of the algorithm's convergence -->
<!-- or in case of having poor initial values the algorithm might diverge altogether (McLachlan & Peel, 2004). -->
<!-- The key is to use a large number of starting value sets. -->
<!-- Geiser (2013) recommends at least 500 sets of initial values in the first optimization step for simple models, -->
<!-- and a greater number of sets for more complex models. -->
<!-- A large number of starting values ensures that we find the true maximum and estimate the model parameters and fit accurately. -->

@@

### Model Fit Indices

@@

LCA studies commonly report -2*log likelihood of the final class solution.
This is a basic fit measure used to compute most information criteria.
However, since log likelihood is not penalized for model complexity,
it will continuously fall with the addition of more classes.

An alternative is using the bootstrapped likelihood ratio test
which can be run using `tidySEM::BLRT()`.
Currently, this test is computationally expensive and can be slow on most computers.
A faster version of this test, 
namely an implementation of *the lazy bootstrap* [@van_kollenburg_lazy_2018]
to `tidySEM` is being developed.

@@

Three important classification diagnostics provided by `tidySEM` are are 
*the minimum* and *maximum percentage of the sample assigned to a particular class*,
*the range of the posterior class probabilities by most likely class membership*, 
and *entropy*. All three are based on posterior class probabilities.
